---
title: 'MIE237'
author: "Neil Montgomery"
date: "2016-01-26"
output: 
  ioslides_presentation: 
    css: 'styles.css' 
    widescreen: true 
    transition: 0.001
---
\newcommand{\Var}[1]{\text{Var}\left( #1 \right)}
\newcommand{\E}[1]{E\left( #1 \right)}
\newcommand{\Sample}[1]{#1_1,\ldots,#1_n}
\newcommand{\od}[2]{\overline #1_{#2\cdot}}
\newcommand{\flist}[2]{\{#1_1, #1_2, \ldots, #1_#2\}}
\newcommand{\samp}[2]{#1_1, #1_2, \ldots, #1_#2}
\renewcommand{\bar}[1]{\overline{#1}}
\newcommand{\ve}{\varepsilon}


```{r, include=FALSE}
knitr::opts_chunk$set(fig.width = 4, fig.height = 4, fig.align = "center")
```

# A taste of categorical data 9.10, 10.13

## Testing independence - more motivation { .build }

100 people (50 male and 50 female) are asked a question about whether they agree or disagree with some political statement. 80 people agreed. 20 people disagreed. Consider the following possible outcomes:

<div class="columns-2">
&nbsp;

Table 1  |  Male  |  Female  | Total
---------|--------|----------|-------
Agree    |   40   |    40    |  80      
Disagree |   10   |    10    |  20     
Total    |   50   |    50    |  100

&nbsp;

Table 2  |  Male  |  Female  | Total
---------|--------|----------|-------
Agree    |   50   |    30    |  80      
Disagree |    0   |    20    |  20     
Total    |   50   |    50    |  100

&nbsp;

Table 3  |  Male  |  Female  | Total
---------|--------|----------|-------
Agree    |   39   |    41    |  80      
Disagree |   11   |     9    |  20     
Total    |   50   |    50    |  100

&nbsp;

Table 4  |  Male  |  Female  | Total
---------|--------|----------|-------
Agree    |   36   |    44    |  80      
Disagree |   14   |     6    |  20     
Total    |   50   |    50    |  100
</div>

## Testing independence

* Treat marginal totals as fixed

* Compute *expected cell counts* $E_i$ assuming independence

* With $O_i$ as *observed cell counts* use the following test statistic:

$$\sum_i \frac{\left(O_i - E_i\right)^2}{E_i} \sim^{\text{approx}} \chi^2_{(r-1)(c-1)}$$

* Approximation is good if $E_i \ge 5$ for all $i$. 

## "more motivation" table calculations 

The table of expected cell counts is:

Expected Cell Counts |  Male  |  Female  | Total
-----|--------|----------|-------
Agree    |   40   |    40    |  80      
Disagree |   10   |    10    |  20     
Total    |   50   |    50    |  100

## p-values { .build }

**Table 1:** $\sum_{i=1}^4 \frac{\left(O_i - E_i\right)^2}{E_i} = 0$ and the p-value is $P\left(\chi^2_1 \ge 0\right) =  `r 1-pchisq(0, 1)`$.

**Table 2:** $\sum_{i=1}^4 \frac{\left(O_i - E_i\right)^2}{E_i} = 10^2/40 + 10^2/40 + 10^2/10 + 10^2/10 = `r sum(10^2/c(40, 40, 10, 10))`$ and the p-value is $P\left(\chi^2_1 \ge 25\right) =  `r round(1-pchisq(sum(10^2/c(40, 40, 10, 10)), 1), 3)`$

**Table 3:** $\sum_{i=1}^4 \frac{\left(O_i - E_i\right)^2}{E_i} = 1^2/40 + 1^2/40 + 1^2/10 + 1^2/10 = `r sum(1^2/c(40, 40, 10, 10))`$ and the p-value is $P\left(\chi^2_1 \ge 0.25\right) =  `r round(1-pchisq(sum(1^2/c(40, 40, 10, 10)), 1), 3)`$

**Table 4:** $\sum_{i=1}^4 \frac{\left(O_i - E_i\right)^2}{E_i} = 4^2/40 + 4^2/40 + 4^2/10 + 4^2/10 = `r sum(4^2/c(40, 40, 10, 10))`$ and the p-value is $P\left(\chi^2_1 \ge 4\right) =  `r round(1-pchisq(sum(4^2/c(40, 40, 10, 10)), 1), 3)`$

## Gas pipe data { .build }

```{r, echo=FALSE, message=FALSE}
library(dplyr)
library(rio)
library(ggplot2)
library(knitr)
library(xtable)

set.seed(1)
n <- 1000
pipeline <- data_frame(Leak = sample(c("Yes", "No"), n, prob = c(0.2,0.8), TRUE),
                       Size = sample(c(1.5, 1.0, 1.75), n, prob=c(0.3, 0.4, 0.3), TRUE),
                       Material = sample(c("Steel", "Aldyl A"), n, TRUE),
                       Pressure = sample(c("Low", "Med", "High"), n, TRUE))
```

Is leak status independent of pipe size? Here is the table summary of the data (leak status in the rows; pipe size in the columns:

`r kable(addmargins(table(pipeline$Leak, pipeline$Size)), )`

The expected cell counts:

`r kable(round(addmargins(chisq.test(pipeline$Leak, pipeline$Size, correct = FALSE)$expected), 1))`

All the $E_i$ easily exceed 5, so the approximation will be good.

## Gas pipe data

The results:

```{r}
chisq.test(pipeline$Leak, pipeline$Size)
```

Not surprising since the columns in the simulated data really were simulated independently. 

# Regression

## Linear models 

>* Model: Output = Input + Noise

>* Linear model: Output = Linear function of inputs + Noise

>* Examples (all with $\varepsilon_i \sim N(0, \sigma^2)$):

>* $Y_i = \mu + \varepsilon_i$

>* $Y_{ij} = \mu_i + \varepsilon_{ij},\quad i\in\{1,2\}$

>* $Y_{ij} = \mu_i + \varepsilon_{ij},\quad i\in\{1,\ldots, k\}$

>* $Y_{i} = \beta_0 + \beta_1 x_i + \varepsilon_{i},\quad x_i\in\{0,1\}$

>* $Y_{i} = \beta_0 + \beta_1 x_i + \varepsilon_{i},\quad x_i\in\mathbb{R}$

>* $Y_{i} = \beta_0 + \beta_1 x_{1i} 
+\beta_2 x_{2i}+ \varepsilon_{i},\quad x_{ji}\in\mathbb{R}$

## Linear models in matrix form

$$\boldsymbol{Y} = \boldsymbol X\boldsymbol{\beta}+\boldsymbol{\varepsilon}$$

<div class="columns-2">
$$\boldsymbol{Y} = \begin{bmatrix}Y_1\\\vdots\\Y_n\end{bmatrix}$$

$$\boldsymbol{\beta} = 
\begin{bmatrix}\beta_0\\\beta_1\\\vdots\\\beta_k
\end{bmatrix}$$

$$\boldsymbol{\varepsilon} = 
\begin{bmatrix}
\varepsilon_1\\
\varepsilon_2\\
\vdots\\
\varepsilon_n\\
\end{bmatrix}$$ 
</div>

## Linear models in matrix form

$$\boldsymbol{Y} = \boldsymbol X\boldsymbol{\beta}+\boldsymbol{\varepsilon}$$

$$\boldsymbol{X} = 
\begin{bmatrix}
1&x_{11}&x_{21}&\cdots&x_{k1}\\
1&x_{12}&x_{22}&\cdots&x_{k2}\\
\vdots&\vdots&\vdots&\vdots&\vdots\\
1&x_{1n}&x_{2n}&\cdots&x_{kn}\\
\end{bmatrix}$$

## "Simple" linear regression

$$Y_i = \beta_0 + \beta_1 x_i + \ve_i$$

$$\ve \sim N(0,\sigma^2)$$

$\beta_1$ is the "slope" parameter
